Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-16 draft.tex
Began to restructure last section

Another observation during the analysis is very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interested, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
%% ------------------------------------
\subsection{Applicability}
%%
Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
%%
Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, making their results harder to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our brushable visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology \cite{Klemm2014VIS}.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.

\subsection{Summary and Future Work}
%%
\noindent We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
%%
Our method made it possible to assess their discriminative power, which is largely limited for separating \emph{male} and \emph{female} subjects, \emph{nutrition} variables, as well as different disease indicators.
%%
The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
%%
Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
%%

>>>>

Another observation during the analysis is that large decision trees achieve low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\paragraph{Applicability.}
%%
Classification methods based on decision trees showed to be useful to assess the discriminative power of a variable set.
%%
Their ability to assess variable combination makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, making their results harder to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our brushable visualization.
%%
Therefore the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows both to assess the discriminative power of the data set regarding the target variables, as well as the most important parameters for the classification.
%Except for gender in the age-plot, no variable could be explained with reasonable accuracy for age or body size.
\paragraph{Predictive Power of Image-Derived Variables.}
We showed, that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not enough to characterize lumbar back pain in the SHIP data set.
%%
Our method made it possible to assess their discriminative power, which is largely limited for separating \emph{male} and \emph{female} subjects, \emph{nutrition} variables, as well as different disease indicators.
%%
The C4.5 algorithm showed to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image parameters.
%%
Over-fitting to the data indicated by complex decision trees have to be taken into account into the analysis as well.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

A very early sign of pathological deformation is the \emph{bone resorption in the center of a vertebra}.
%%
They change from being block-shaped to be dented in the center (Fig~\ref{fig:proposedFeatures}~a).
%%
This information can be obtained by segmenting the whole vertebra or the top and bottom point if each vertebra center.
%%
Another valuable variable would be the \emph{spine canal thickness} (Fig~\ref{fig:proposedFeatures}~c).
%%
Low spine canal thickness can be an indicator for an impending herniated disk.
%%
Both surface texture of the vertebra and thickness of the spine canal are used to diagnose herniated disks.
%%
The \emph{overall spine canal shape} is also of interest, since scoliosis and lordosis can be characterized more precise by deriving the Cobb angles from this shape (Fig~\ref{fig:proposedFeatures}~b,~c).
%% ------------------------------------
\subsection{Future Work}
%% Contributions
We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
%%
We applied the technique to derive insight into the predictive power of 9 image-derived variables for 134 non-image variables.
%%TODO HIER WEITER
%We carried out an analysis on subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine.
%To assess the influence of parameters which strongly influence the spine shape, we applied the technique to.
%%
